---
title: "How to use EPos-MoL"
author: "Nils Hoffmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
renv::install(c("tidyverse","readxl","lifs-tools/rgoslin","here"))
library(tidyverse)
library(readxl)
library(rgoslin)
library(here)
```

```{r, echo=TRUE}
scoringScheme <- read_excel(
  path = here("inst","extdata","Table 1.xlsx"),
  sheet = "Scoring scheme", 
  range = "A1:X30"
)
```

```{r}
scoringSchemeLong <- scoringScheme %>% 
  pivot_longer(
    cols=!Primary:ID, 
    names_to = "lipidClassification"
  )
scoringSchemeLong
```

## Example a

|               |                     |
| ------------- | ------------------- |
| Platform      | TQ EVOQ Elite       |
| Separation    | UHPLC               |
| MS Mode       | MRM                 |
| Ion Mode      | (+)                 |

```{r}
ex_a <- "Cer 18:1;O2/16:0"
ex_a_norm <- parseLipidNames(ex_a)
ex_a_norm
```

```{r}
ex_a_lc <- ex_a_norm[1, "Lipid.Maps.Category"]
nominalMassPos <- scoringSchemeLong |> 
  filter(
    Primary=="MS1" & 
    Secondary=="Molecular properties" &
    Evidence=="Nominal mass" &  
    lipidClassification==ex_a_lc
  )|> 
  select(value)
separation <- scoringSchemeLong |> 
  filter(
    Primary=="MS1" &
    Secondary=="Molecular properties" &  
    Evidence=="Chromatography (RT)" &  
    lipidClassification==ex_a_lc
  ) |> 
  select(value)
headGroup <- scoringSchemeLong |> 
  filter(
    Primary=="MS2" &
    Secondary=="CID fragments" &  
    Fragment=="LCF" &  
    Evidence=="Headgroup 1" &  
    lipidClassification==ex_a_lc
  ) |> 
  select(value)
score <- sum(  
  nominalMassPos,
  separation,
  headGroup
)                                                                        
score
```
## Example b

Example b assigns points for identification of a Ceramide with one long chain base and a fatty acyl, both with 18 carbon atoms and one double bond each. Identification was performed in positive and negative mode on a low resolution QTrap instrument, without prior separation. We thus assign points for nominal mass accuracy in positive and negative mode, additional CID fragments for MS2 in both modes and the primary fatty acyl fragment for negative mode.

|               |                     |
| ------------- | ------------------- |
| Platform      | QTrap 6500          |
| Separation    | NA                  |
| MS Mode       | Precursor Ion Scan  |
| Ion Mode      | (+) & (-)           |
```{r}
ex_b <- "Cer 18:1;O2/18:1"
ex_b_norm <- parseLipidNames(ex_b)
ex_b_norm
```

```{r}
ex_b_lipidCategory <- ex_b_norm[1, "Lipid.Maps.Category"]
nominalMassPos <- scoringSchemeLong |> 
  filter(
    Primary=="MS1" & 
    Secondary=="Molecular properties" &
    Evidence=="Nominal mass" &  
    lipidClassification==ex_b_lipidCategory
  ) |> 
  select(value)
headGroup1Pos <- scoringSchemeLong |> 
  filter(
    Primary=="MS2" &
    Secondary=="CID fragments" &  
    Fragment=="LCF" &  
    Evidence=="Headgroup 1" &  
    lipidClassification==ex_b_lipidCategory
  ) |> 
  select(value)
nominalMassNeg <- scoringSchemeLong |> 
  filter(
    Primary=="MS1" & 
    Secondary=="Molecular properties" &
    Evidence=="Nominal mass" &  
    lipidClassification==ex_b_lipidCategory
  ) |> 
  select(value)
headGroup1Neg <- scoringSchemeLong |> 
  filter(
    Primary=="MS2" &
    Secondary=="CID fragments" &  
    Fragment=="LCF" &  
    Evidence=="Headgroup 2" &  
    lipidClassification==ex_b_lipidCategory
  ) |>
  select(value)
fa1Neg <- scoringSchemeLong |> 
  filter(
    Primary=="MS2" &
    Secondary=="CID fragments" &  
    Fragment=="MLF" &  
    Evidence=="Fatty acyl 1-1" &  
    lipidClassification==ex_b_lipidCategory
  ) |>
  select(value)
score <- sum(  
  nominalMassPos,
  headGroup1Pos,
  nominalMassNeg,
  headGroup1Neg,
  fa1Neg
)                                                                        
score
```
